---
title: "Week 5 Assignment"
author: "Xhulia Turkaj"
date: "2023-10-01"
output: html_document
---

## Create a .CSV file (or optionally, a MySQL database!) that includes all of the information above.
You’re encouraged to use a “wide” structure so that you can practice tidying and transformations. 

# Load the libraries 

```{r}
# Load the libraries 
library(tidyverse)
library(dplyr)
library(ggplot2)
library(scales)
```


#Create the dataframe

```{r}



flights <- rbind(c("Airline", "Status", "Los Angeles", "Phoenix", "San Diego", "San Francisco", "Seattle"),
             c("ALASKA", "on time", 497, 221, 212, 503, 1841),
             c(NA, "delayed", 62, 12, 20, 102, 305),
             c("AM WEST", "on time", 694, 4840, 383, 320, 201),
             c(NA, "delayed", 117, 415, 65, 129, 61))

write.table(flights, file = "week5flights.csv", sep = ",", col.names=F, row.names=F)

```

# Read the csv file from github
```{r}
#Read the csv file 

airlines <- read.csv("https://raw.githubusercontent.com/xhuliaturkaj/week_5/main/week5flights.csv")

airlines
```

## Structuring data 

```{r}
 # First start by filling the Airline Name in the rows where it is missing
airlines <- airlines %>%
  fill(Airline, .direction = "down")
airlines
```

```{r}
#Next use pivot longer to reshape the dataframe
airlines_long <- airlines %>%
  pivot_longer(cols = -c(Airline, Status), names_to = "City", values_to = "Count")
airlines_long
```
##  Perform analysis to compare the arrival delays for the two airlines.

Let's begin by investigating the delays for both airlines
```{r}
# I will summarize the Status counts grouping by Airline

delay_airlines <- airlines_long %>%
  group_by(Airline, Status) %>%
  summarize(Total = sum(Count, na.rm = TRUE)) %>%
  mutate(Proportion = Total / sum(Total))

print(delay_airlines)

```
Next I will visualize the information above 

```{r}

ggplot(delay_airlines, aes(x = Airline, y = Proportion, fill = Status)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.9)) +
  geom_text(aes(label = percent(Proportion)), 
            position = position_dodge(width = 0.9), vjust = -0.5) +
  labs(y = "Proportion of Flights", 
       title = "Proportion of On-time and Delayed Flights for Each Airline") +
  scale_fill_manual(values = c("on time" = "green", "delayed" = "red"))
```



#### The comparison indicates that ALASKA  airline has an 86.7% on-time arrival rate and a 13.3% delay rate, whereas AM WEST airline demonstrates a higher on-time arrival rate of 89.1% and a lower delay rate of 10.9%, suggesting AM WEST is relatively more punctual in its services.


Next I would like to explore how these rates vary by city. 

```{r}

city_proportions <- airlines_long %>%
  group_by(Airline, City, Status) %>%
  summarise(Count = sum(Count, na.rm = TRUE), .groups = 'drop') %>%
  group_by(Airline, City) %>%
  mutate(Total = sum(Count), Proportion = Count / Total) %>%
  ungroup()

city_proportions

```
Let's use  city_proportions that I created above to visualize these proportions 

```{r}
ggplot(city_proportions, aes(x = City, y = Proportion, fill = Status)) +
  geom_bar(stat = "identity", position = "dodge", show.legend = TRUE) +
  facet_wrap(~Airline, scales = "free_y") +
  labs(y = "Proportion of Flights",
       x = "City",
       title = "Proportion of On-time and Delayed Flights for Each Airline by City") +
  scale_fill_manual(values = c("on time" = "green", "delayed" = "red")) +
  geom_text(aes(label = scales::percent(Proportion, accuracy = 1)),
            position = position_dodge(width = 0.9), vjust = -0.25) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Conclusion

#### As we can see from the visual representation there are significant differences in the delay/on-time arrival  rates between the two airlines in several cities, with AM WEST typically showing a greater percentage of flights arriving on time, particularly in Phoenix with 95% (on time arrivals) , while Alaska Airlines performs exceptionally well in locations such as Phoenix (92% on time arrivals) and San Diego (85% on time arrivals). San Francisco and Seattle seemed to be cities with a high arrival delay rate for both airlines but specifically for AM WEST. A statistical analysis is essential  to determine whether the observed differences in arrival times are statistically significant.

